Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-16 iso-int/rc-iso-int.tex
redrew the tree diagrams
shorter in pages now
space fitting
_
small corrections

The branch labeled {{((), m)}} corresponds to the machine state {{<0,m>}} and
hence looks like the stop state of the machine. Similarly the branch labeled
{{((), n')}} corresponds to the start state of the machine. We are close to
the completion of the interpreter at this point. The last step in the
construction involves introducing a {{trace}] operation for iteration of
reduction steps and exposing the start and stop states. We do this by
``sliding'' the two sections of the diagram as shown below:

@@

A closer look however reveals that the machine only defines a \emph{partial}
isomorphism: not all valid final states can be mapped to valid start
states. Indeed consider the configuration {{<0,2,3>}} which is a valid final
state. Going backwards, the transitions start as follows
{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine 
gets stuck at a state that is not a valid start state. 
There are several ways to make this machine reversible. The simplest is to
add another possible kind of start state {{<n,0,n>}}. Our construction now
proceeds as expected. After expanding the input to expose $n$ and the output
to expose both $p$ and $m$, we reach the following configuration:

@@

The type of binary trees we use is {{@@x.nat+x*x}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
subtree is currently being explored. The definitions are:

%subcode{bnf} include main
% Tree, t = L n | N t t
% Tree Contexts, c = Lft c t | Rgt t c | []
%
% Machine states = <t, c> | {[c, t]}
% Start state = <t, []>
% Stop State = {[[], t]}

Our example machine below traverses a given tree and increments every leaf
value:

@@

The machine here is a little richer than the ones dealt with
previously. In particular, we have a two types of machine states --
{{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds to walking
down a tree, building up the context in the process. The second
corresponds to reconstructing the tree from the context and also
switching focus to any unexplored subtrees in the process. There are
also two syntactic categories to deal with (trees and tree contexts)
where previously we only had numbers. The {{fold}} and {{unfold}}
isomorphisms that we need for trees and tree contexts are:

@@

%subcode{opsem} include main
%! columnStyle = rclr
% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4

And so machine states have the form:

%subcode{bnf} include main
% Machine states = <bool, t, c>
% Start state = <T, t, []>
% Stop state =  <F, t, []>

@@

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/tree1.pdf}
}
\end{center}

@@

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/tree2.pdf}
}
\end{center}

@@

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi0-1.pdf}
}
\end{center}

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi0-2.pdf}
}
\end{center}

>>>>

The branch labeled {{((), m)}} corresponds to the machine state {{<0,m>}} and
hence looks like the stop state of the machine. Similarly the branch labeled
{{((), n')}} corresponds to the start state of the machine. We are close to
the completion of the interpreter at this point. The last step in the
construction involves introducing a {{trace}} operation for iteration of
reduction steps and exposing the start and stop states. We do this by
``sliding'' the two sections of the diagram as shown below:

@@

A closer look however reveals that the machine defines a \emph{partial}
isomorphism: not all valid final states can be mapped to valid start
states. Indeed consider the configuration {{<0,2,3>}} which is a valid final
state. Going backwards, the transitions start as follows 
{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine gets stuck 
at a state that is not a valid start state.  There are several ways to deal with this
situation. The first is to use a partial isomorphism that diverges on the stuck state.
The second is to add another
possible kind of start state {{<n,0,n>}} to make the isomorphism total. 
Assuming the seond solution, our construction now proceeds as
expected. After expanding the input to expose $n$ and the output to expose
both $p$ and $m$, we reach the following configuration:

@@

The type of binary trees we use is {{@@x.nat+x*x}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
subtree is currently being explored. The definitions are shown on the left:

\begin{multicols}{2}
%subcode{bnf} include main
% Tree, t = L n | N t t
% Tree Contexts, c = Lft c t | Rgt t c | []
%
% Machine states = <t, c> | {[c, t]}
% Start state = <t, []>
% Stop State = {[[], t]}

@@

\end{multicols}

The reduction rules on the right traverses a given tree and increments
every leaf value.  The machine here is a little richer than the ones
dealt with previously. In particular, we have a two types of machine
states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
to walking down a tree, building up the context in the process. The
second corresponds to reconstructing the tree from the context and
also switching focus to any unexplored subtrees in the process. There
are also two syntactic categories to deal with (trees and tree
contexts) where previously we only had numbers. The {{fold}} and
{{unfold}} isomorphisms that we need for trees and tree contexts are:

@@

\begin{multicols}{2}
%subcode{bnf} include main
% Machine states = <bool, t, c>
%
% Start state = <T, t, []>
% Stop state =  <F, t, []>

%subcode{opsem} include main
%! columnStyle = rclr
% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
\end{multicols}

@@

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/tree1.pdf}
}
\end{center}

@@

\begin{center}
\scalebox{1.1}{
  \includegraphics{diagrams/tree2.pdf}
}
\end{center}

@@

%% \begin{center}
%% \scalebox{0.85}{
%%   \includegraphics{diagrams/pi0-1.pdf}
%% }
%% \end{center}

\begin{center}
\scalebox{0.85}{
  \includegraphics{diagrams/pi0-2.pdf}
}
\end{center}
